Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-09-12 Paper/SFT.tex
polish

    \path[->] 
    (q_0) edge node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_1) edge node[right] {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2)
    (q_1) edge[loop above] node {[$'\texttt{a}'-'\texttt{z}'$],$f_1$} (q_1)
    (q_2) edge[loop below] node {[$'\texttt{A}'-'\texttt{Z}'$],$f_2$} (q_2);
  \end{tikzpicture}
    \caption{An example of SFT}
    \label{fig-example-ft}
\end{figure}

    Figure \ref{fig-example-ft} illustrates an SFT that accepts words matching the regular expression $/['\texttt{a}'-'\texttt{z}']+['\texttt{A}'-'\texttt{Z}']+/$. The transition labels utilize intervals of the form $[i\text{-}j]$, which are interpreted as predicates $\lambda x.~i \leq x \leq j$. The output functions $f_1$ and $f_2$ perform case transformations: $f_1 = \lambda x.~\texttt{toUpper}(x)$ converts lowercase letters to uppercase, while $f_2 = \lambda x.~\texttt{toLower}(x)$ performs the inverse operation. For example, given the input string \texttt{"bigSMALL"}, this SFT produces the output \texttt{"BIGsmall"}.

>>>>

    \path[->] 
    (q_0) edge node {[$\texttt{a}-\texttt{z}$],$f_1$} (q_1)
    (q_1) edge node[right] {[$\texttt{A}-\texttt{Z}$],$f_2$} (q_2)
    (q_1) edge[loop above] node {[$\texttt{a}-\texttt{z}$],$f_1$} (q_1)
    (q_2) edge[loop below] node {[$\texttt{A}-\texttt{Z}$],$f_2$} (q_2);
  \end{tikzpicture}
    \caption{An example of SFT}
    \label{fig-example-ft}
\end{figure}

Figure \ref{fig-example-ft} illustrates an SFT that accepts words matching the regular expression $/[\texttt{a}-\texttt{z}]+[\texttt{A}-\texttt{Z}]+/$. The transition labels utilize intervals of the form $[i\text{-}j]$, which are interpreted as predicates $\lambda x.~i \leq x \leq j$. The output functions $f_1$ and $f_2$ perform case transformations: $f_1 = \lambda x.~\texttt{toUpper}(x)$ converts lowercase letters to uppercase, while $f_2 = \lambda x.~\texttt{toLower}(x)$ performs the inverse operation. For example, given the input string \texttt{bigSMALL}, this SFT produces the output \texttt{BIGsmall}.

### 2025-09-12 Paper/application.tex
polish

The string replacement, denoted as
$\texttt{replace}(\texttt{str}, \texttt{pattern}, \newline\texttt{replacement}, \texttt{flag})$, is a fundamental string transformation accepting four parameters:
(1) $\texttt{str}$: The input string to be transformed.
(2) $\texttt{pattern}$: A regular expression defining the matching criteria.
(3) $\texttt{replacement}$: The string that replaces matched substrings.
(4) $\texttt{flag}=s/g$: The replacement mode flag, where $s$ denotes single first occurrence replacement and $g$ denotes all occurrences replace operation.

@@

To illustrate how SFTs model this operation, consider $\texttt{replace}(s, \texttt{/[0-9]+/}, "\texttt{NUM}", \texttt{flag})$. We construct the corresponding SFT by the following three steps.

\begin{enumerate}
  \item Construct an SFA recognizing the pattern $\texttt{/[0-9]+/}$ and transform it into an SFT by augmenting each transition with output function "$f = \lambda x.~\texttt{None}$" that produces the empty string.
  \item Construct an SFA accepting the replacement string "\texttt{NUM}". Since constant strings are specialized regular expressions, we build their SFA representation and convert it to an SFT by adding $\varepsilon$-transitions with output functions that emit the replacement characters sequentially.
  \item Compose the two SFTs by connecting the accepting state of the pattern-matching SFT to the initial state of the replacement-generating SFT via $\varepsilon$-transitions.
\end{enumerate}

@@

Figure \ref{fig-rearranged-automata} presents the complete SFT construction, obtained by composing the pattern-matching SFT generated from \texttt{/[0-9]+/}  with the replacement-generating SFT generated from "\texttt{NUM}", where the function \texttt{f} is defined as "$f = \lambda x.~\texttt{None}$" and the function \texttt{g} is defined as:
\[
\texttt{g} = \lambda~i~x.~\texttt{match}~i~\texttt{with}~
\begin{cases}
1 \mapsto [(78, 78)] & \text{(ASCII for 'N')} \\
2 \mapsto [(85, 85)] & \text{(ASCII for 'U')} \\
3 \mapsto [(77, 77)] & \text{(ASCII for 'M')} \\
\_ \mapsto \texttt{None}
\end{cases}
\]

@@

     \path[->] 
     (q0') edge [bend left] node {\texttt{[0-9]}, $\texttt{f}$} (q1')
     (q1') edge [loop right] node {\texttt{[0-9]}, $\texttt{f}$} (q1');

@@

    \path[->] 
    (q0'') edge node {\texttt{None}, \texttt{(g 1)}} (q1'')
    (q1'') edge [bend left=5] node[left] {\texttt{None}, \texttt{(g 2)}} (q2'')
    (q2'') edge node {\texttt{None}, \texttt{(g 3)}} (q3'')
    (q1') edge [bend right=30] node {\texttt{None}, \texttt{f}} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation \\ $\texttt{replace}(s, \texttt{/[0-9]+/}, \text{``NUM''})$}
  \label{fig-rearranged-automata}
  \end{figure}

@@

We evaluate our SFT-based replacement operation by extending CertiStr \cite{cpp/KanLRS22}, a certified string solver that supports only basic string operations such as concatenation and regular constraints. CertiStr employs a forward-propagation approach to solve string constraints, which requires computing the forward image of string operations. Our SFT product operation provides an exact forward image computation for replacement operations, enabling seamless integration of replacement functionality into CertiStr's solving framework.

@@

We also compare our approach with two state-of-the-art string solvers: Ostrich \cite{pacmpl/ChenFHHHKLRW22} and CVC5 \cite{cvc5}.
The experimental results demonstrate that string replacement operations, particularly replace-all operations, pose significant challenges for existing state-of-the-art solvers.
The "unknown" label in the table indicates instances where CVC5 cannot determine satisfiability. 
Notably, PCP and RNA benchmarks cannot be solved by most state-of-the-art string
solvers, including CVC5 \cite{cvc5} and Z3Str \cite{Z3-str3}, yet CertiStrR solves
them efficiently. Among existing solvers, only OSTRICH \cite{pacmpl/ChenFHHHKLRW22} can handle these benchmarks.
CertiStrR outperforms both Ostrich and CVC5 across most benchmark subgroups, with the only exception of the "Replace Str" category where CVC5 achieves better performance (though it experiences one timeout). CertiStrR demonstrates particularly strong performance in the "Replace All" category.
These results demonstrate the effectiveness and efficiency of our certified SFT formalization and implementation.

@@

Consider the transition $[\texttt{a-z}]/id$, where $id$ is the identity function. To model this transition with replacement operations, the interval must be explicitly expanded into the individual characters $\texttt{'a'}$ through $\texttt{'z'}$, requiring 26 nested \texttt{replace-all} operations—one for each character. Thus, modeling $[\texttt{a-z}]/id$ with replacement operations yields an expression of the form:
\[
\texttt{replace}(\texttt{replace}(\ldots,~\texttt{"b"},~\texttt{"b"},~g),~\texttt{"a"},~\texttt{"a"},~g).
\]

>>>>

The string replacement, denoted as
$\texttt{replace}(\texttt{str}, \texttt{pattern}, \newline\texttt{replacement}, \texttt{flag})$, is a fundamental string transformation accepting four parameters:
(1) $\texttt{str}$: The input string to be transformed.
(2) $\texttt{pattern}$: A regular expression defining the matching pattern.
(3) $\texttt{replacement}$: The string that replaces matched substrings.
(4) $\texttt{flag}=s/g$: The replacement mode flag, where $s$ denotes single first occurrence replacement and $g$ denotes all occurrences replacement operation.

@@

To illustrate how SFTs model this operation, consider $\texttt{replace}(s, \texttt{/[0-9]+/},~$\texttt{NUM}$,~\texttt{flag})$. We construct the corresponding SFT by the following three steps.

\begin{enumerate}
  \item Construct an SFA recognizing the pattern $\texttt{/[0-9]+/}$ and transform it into an SFT by augmenting each transition with output function "$\lambda x.~\texttt{None}$" that produces the empty string.
  \item Construct an SFA accepting the replacement string \texttt{NUM} and convert it to an SFT by
  replacing each transtion in the SFA $(q, c, q')$ with
  $(q, (\texttt{None},~\lambda \_.~c ), q')$.
  \item Compose the two SFTs by connecting the accepting state of the pattern-matching SFT to the initial state of the replacement-generating SFT via $\varepsilon$-transitions.
\end{enumerate}

@@

Figure \ref{fig-rearranged-automata} presents the complete SFT construction, obtained by composing the pattern-matching SFT generated from \texttt{/[0-9]+/}  with the replacement-generating SFT generated from \texttt{NUM}.

@@

     \path[->] 
     (q0') edge [bend left] node {\texttt{[0-9]}, $\lambda~x.~\texttt{None}$} (q1')
     (q1') edge [loop right] node {\texttt{[0-9]}, $\lambda~x.~\texttt{None}$} (q1');

@@

    \path[->] 
    (q0'') edge node {\texttt{None}, $\lambda\_.\texttt{N}$} (q1'')
    (q1'') edge [bend left=5] node[left] {\texttt{None}, $\lambda\_.\texttt{U}$} (q2'')
    (q2'') edge node {\texttt{None}, $\lambda\_.\texttt{M}$} (q3'')
    (q1') edge [bend right=30] node {\texttt{None}, $\lambda x.~\texttt{None}$} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation $\texttt{replace}(s,\texttt{/[0-9]+/}, \texttt{NUM})$}
  \label{fig-rearranged-automata}
  \end{figure}

@@

We evaluate our SFT-based replacement operation by extending CertiStr \cite{cpp/KanLRS22}, a certified string solver that previously supported only basic string operations such as concatenation and regular constraints. The original CertiStr was incapable of verifying any XSS attack due to the lack of support for transducers and replacement operations.
CertiStr employs a forward-propagation approach to solve string constraints, which requires computing the forward image of string operations. Our SFT product operation provides an exact forward image computation for replacement operations, enabling seamless integration of replacement functionality into CertiStr's solving framework.

@@

We also compare our approach with two state-of-the-art string solvers: Ostrich \cite{pacmpl/ChenFHHHKLRW22} and CVC5 \cite{cvc5}.
The experimental results demonstrate that string replacement operations, particularly replace-all operations, pose significant challenges for existing state-of-the-art solvers.
The "unknown" label in the table indicates instances where CVC5 cannot determine satisfiability. 
Notably, PCP and RNA benchmarks cannot be solved by most state-of-the-art string
solvers, including CVC5 \cite{cvc5}, yet CertiStrR solves
them efficiently. Among existing solvers, only OSTRICH \cite{pacmpl/ChenFHHHKLRW22} can handle these benchmarks.
CertiStrR outperforms both Ostrich and CVC5 across most benchmark subgroups, with the only exception of the "Replace Str" category where CVC5 achieves better performance (though it experiences one timeout). CertiStrR demonstrates particularly strong performance in the "Replace All" category.
These results demonstrate the effectiveness and efficiency of our certified SFT formalization and implementation.

@@

Consider the transition $[\texttt{a-z}]/id$, where $id$ is the identity function. To model this transition with replacement operations, the interval must be explicitly expanded into the individual characters $\texttt{a}$ through $\texttt{z}$, requiring 26 nested \texttt{replace-all} operations—one for each character. Thus, modeling $[\texttt{a-z}]/id$ with replacement operations yields an expression of the form:
\[
\texttt{replace}(\texttt{replace}(\ldots,~\texttt{b},~\texttt{b},~g),~\texttt{a},~\texttt{a},~g).
\]

### 2025-09-12 Paper/formalization.tex
polish

We present the SFT formalization in abstract layer in Fig. \ref{fig-def-FT}. While the elements $\mathcal{Q}_t$, $\mathcal{I}_t$, and $\mathcal{F}_t$ directly correspond to their counterparts ($\mathcal{Q}$, $\mathcal{I}$, and $\mathcal{F}$) in Definition \ref{def-sft}, the transition relations are not exactly the same.
%
The transition relations are formalized through \texttt{LTTS} (Labeled Transducer Transition System), where each transition is represented as a triple $'q \times ('a \texttt{ set option }\times 'i) \times 'q$. This representation reflects several key design decisions aimed at enhancing the abstraction and flexibility of our SFT formalization.
\begin{figure}[hbt!]
	\begin{lstlisting}
record (|$'q,$||$~'a,$| |$'i,$| |$'b$|) |$\texttt{NFT}$| =
	|$\mathcal{Q}_t$| :: "|$'q$| set"
	|$\Delta_t$| :: "(|$'q,$||$~'a,$| |$'i$|) LTTS"
	|$\mathcal{I}_t$| :: "|$'q$| set"
	|$\mathcal{F}_t$| :: "|$'q$| set"
        |$\mathcal{M}_t$| :: "|$'i\Rightarrow ('a,~'b)$| Tlabel"
type_synonym |$('q,~'a,~'i)$| LTTS = "|$('q \times ('a \text{ set option }\times 'i) \times 'q)$| set"
type_synonym |$('a,~'b)$| Tlabel = "|$'a \text{ option}~\Rightarrow~'b$ set option"
	\end{lstlisting}
\caption{The formalization of $SFTs$ in Isabelle/HOL}
\label{fig-def-FT}
\end{figure}

@@

Figure \ref{fig-bool-algebra-locale} presents the simplified definition of a boolean algebra locale. The type variable \texttt{'b} represents any boolean algebra, while the type variable \texttt{'a} denotes the element type of the sets that the boolean algebra represents. The locale defines a collection of operations and assumptions that characterize boolean algebra behavior, including operations for set semantics, emptiness checks, non-emptiness checks, intersection, difference, and element membership. The function "\texttt{sem}" provides the semantic interpretation of a boolean algebra as a set, with all other operations' assumptions defined in terms of this set semantics.

The locale serves as an interface to boolean algebra implementations. Any concrete boolean algebra implementation must provide a concrete instantiation of this locale, which includes defining the type variable \texttt{'b} and providing implementations for all operations along with proofs of the required assumptions.

\begin{figure}[hbt!]
\begin{lstlisting}[mathescape=true]
locale bool_algebra =
 fixes sem :: " 'b $\Rightarrow$ 'a::ord set"
 fixes empty:: "'b $\Rightarrow$ bool"
 fixes nempty:: "'b $\Rightarrow$ bool"
 fixes intersect:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes diff:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes elem:: "'a  $\Rightarrow$ 'b $\Rightarrow$ bool"
 assumes
  empty_sem: "empty s = (sem s = {})"
                                    and
  nempty_sem: "nempty s = $\neg$ (empty s)" 
                                    and
  inter_sem: "sem (intersect s1 s2) =
               (sem s1) $\cap$ (sem s2)" and
  diff_sem: "sem (diff f1 f2 s1 s2) =
               (sem s1) - (sem s2)" and
  elem_sem: "elem a s $\equiv$  (a $\in$ sem s)"
\end{lstlisting}
\caption{The boolean algebra locale}
\label{fig-bool-algebra-locale}
\end{figure}

@@

Figure \ref{fig-compute-nft-product} illustrates the core algorithm \texttt{productT\_impl}, which is implemented using the \emph{Refine\_Monadic} framework. Note that in the implementation, there are some refined operations: \texttt {nft\_tranfun}, \texttt{nfa\_states}, \texttt{nfa\_trans}, \texttt{nfa\_initial}, and \texttt{nfa\_accepting}. These are corresponding to the states, transitions, initial states, accepting states, and output function mapping of the SFT.

>>>>

We present the SFT formalization in abstract layer in Fig. \ref{fig-def-FT}. While the elements $\mathcal{Q}_t$, $\mathcal{I}_t$, and $\mathcal{F}_t$ directly correspond to their counterparts ($\mathcal{Q}$, $\mathcal{I}$, and $\mathcal{F}$) in Definition \ref{def-sft}, the transition relations are not exactly the same.
%
The transition relations are formalized through \texttt{LTTS} (Labeled Transducer Transition System), where each transition is represented as a triple $'q \times ('a \texttt{ set option }\times 'i) \times 'q$. This representation reflects several key design decisions aimed at enhancing the abstraction and flexibility of our SFT formalization.
\begin{figure}[hbt!]
	\begin{lstlisting}
record (|$'q,$||$~'a,$| |$'i,$| |$'b$|) |$\texttt{NFT}$| =
	|$\mathcal{Q}_t$| :: "|$'q$| set"
	|$\Delta_t$| :: "(|$'q,$||$~'a,$| |$'i$|) LTTS"
	|$\mathcal{I}_t$| :: "|$'q$| set"
	|$\mathcal{F}_t$| :: "|$'q$| set"
        |$\mathcal{M}_t$| :: "|$'i\Rightarrow ('a,~'b)$| Tlabel"
type_synonym |$('q,~'a,~'i)$| LTTS = "|$('q \times ('a \text{ set option }\times 'i) \times 'q)$| set"
type_synonym |$('a,~'b)$| Tlabel = "|$'a \text{ option}~\Rightarrow~'b$ set option"
	\end{lstlisting}
\caption{The formalization of SFTs in Isabelle/HOL}
\label{fig-def-FT}
\end{figure}

@@

Figure \ref{fig-bool-algebra-locale} presents the simplified definition of a boolean algebra locale. The type variable \texttt{$'$b} represents any boolean algebra, while the type variable \texttt{$'$a} denotes the element type of the sets that the boolean algebra represents. The locale defines a collection of operations and assumptions that characterize boolean algebra behavior, including operations for set semantics, emptiness checks, non-emptiness checks, intersection, difference, and element membership. The function "\texttt{sem}" provides the semantic interpretation of a boolean algebra as a set, with all other operations' assumptions defined in terms of this set semantics.

The locale serves as an interface to boolean algebra implementations. Any concrete boolean algebra implementation must provide a concrete instantiation of this locale, which includes defining the type variable $'$\texttt{b} and providing implementations for all operations along with proofs of the required assumptions.

\begin{figure}[hbt!]
\begin{lstlisting}[mathescape=true]
locale bool_algebra =
 fixes sem :: " 'b $\Rightarrow$ 'a::ord set"
 fixes empty:: "'b $\Rightarrow$ bool"
 fixes nempty:: "'b $\Rightarrow$ bool"
 fixes intersect:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes diff:: "'b $\Rightarrow$ 'b $\Rightarrow$ 'b"
 fixes elem:: "'a  $\Rightarrow$ 'b $\Rightarrow$ bool"
 assumes
  empty_sem: "empty s = (sem s = {})"
                                    and
  nempty_sem: "nempty s = $\neg$ (empty s)" 
                                    and
  inter_sem: "sem (intersect s1 s2) =
               (sem s1) $\cap$ (sem s2)" and
  diff_sem: "sem (diff s1 s2) =
               (sem s1) - (sem s2)" and
  elem_sem: "elem a s $\equiv$  (a $\in$ sem s)"
\end{lstlisting}
\caption{The boolean algebra locale}
\label{fig-bool-algebra-locale}
\end{figure}

@@

Figure \ref{fig-compute-nft-product} illustrates the algorithm \texttt{productT\_impl}, which is implemented using the \emph{Refine\_Monadic} framework. Note that in the implementation, there are some refined operations:  \texttt{nfa\_states}, \texttt{nfa\_trans}, \texttt{nfa\_initial}, \texttt{nfa\_accepting},
and \texttt{nft\_tranfun}. These are corresponding to the states, transitions, initial states, accepting states, and output function mapping of the SFT.

### 2025-09-12 Paper/introduction.tex
polish

To evaluate the effectiveness and efficiency of our formalization, we 
provide two applications: (1) analysis of the web application sanitizers for html, css, json, and javascript
which are used for preventing XSS attacks, and (2) an SMT-LIB \cite{smtlib}
compliant certified string solver handling equality, concatenation, regular
constraints, and more importantly the string replaceall operator, which has been
argued to be indispensable for analysis of XSS 
\cite{DBLP:conf/popl/LinB16,Kern,Berkeley-JavaScript,systematic-transduction,uss/HooimeijerLMSV11}.
The experimental results demonstrate that our formalization achieves computational efficiency in constraint-solving scenarios.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers (SFTs) in Isabelle/HOL.
We prove the correctness of key operations of SFTs and symbolic SFAs, such as the product operation between SFTs and SFAs.
\item We leverage the refinement framework to develop an extensible and efficient SFT formalization capable of supporting diverse Boolean algebras, ensuring adaptability to future transition label representations. The correctness of the refined, efficient implementation is proved via structural isomorphism between refinement layers.
\item We develop a certified interval algebra with verified set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization by applying it to sanitizer analysis and integrating our formalized SFT into the verified forward-propagation string solver CertiStr \cite{cpp/KanLRS22}. This integration extends CertiStr from supporting only simple concatenation and regular constraints to handling more sophisticated string replacement operations.
\end{enumerate}

>>>>

To evaluate the effectiveness and efficiency of our formalization, we 
provide two applications: (1) analysis of the web application sanitizers for html, css, json, and javascript
which are used for preventing XSS attacks, and (2) a certified string solver handling equality, concatenation, regular
constraints, and more importantly the string replaceall operator, which has been
argued to be indispensable for analysis of XSS
\cite{DBLP:conf/popl/LinB16,Kern,Berkeley-JavaScript,systematic-transduction,uss/HooimeijerLMSV11}.
The experimental results demonstrate that our formalization achieves computational efficiency in constraint-solving scenarios.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers  in Isabelle/HOL.
We prove the correctness of key operations of SFTs and symbolic SFAs, such as the product operation between SFTs and SFAs.
\item We leverage the refinement framework to develop an extensible and efficient SFT formalization capable of supporting diverse Boolean algebras, ensuring adaptability to future transition label representations. The correctness of the refined, efficient implementation is proved via structural isomorphism between refinement layers.
\item We develop a certified interval algebra with verified set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization by applying it to sanitizer analysis and integrating our formalized SFT into the verified forward-propagation string solver CertiStr \cite{cpp/KanLRS22}. This integration extends CertiStr from supporting only simple concatenation and regular constraints to handling more sophisticated string replacement operations.
\end{enumerate}
